---
title: "Reproducible Research - Assignment 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Task 1 - Load the data & plot the histogram of the total number of steps taken each day


```{r load data}
setwd("C:/Users/Siddharth/Documents/Careers/Siddharth/DataScienceSpecialization/Reproducible Research/Week 2/")

activity <- read.csv("activity.csv")
activity <- na.omit(activity)

total_steps <- tapply(activity$steps, activity$date, sum)

total_steps <- total_steps[!is.na(total_steps)]

hist(total_steps, col = "red", xlab = "Total number of steps taken each day", main = "Histogram of steps taken each day")

mean_total_steps <- as.integer(mean (total_steps))

median_total_steps <- median (total_steps)


```

The mean for the number of steps taken each day is `r mean_total_steps`

The median for the number of steps taken each day is `r median_total_steps`

## Task 2 - Find the average daily pattern

Calculate the average number of steps taken for each time interval across all the days

```{r average daily pattern}

mean_steps <- tapply(activity$steps, activity$interval, mean)

mean_steps_df <- data.frame(mean_steps)

mean_steps_df2 <- cbind(mean_steps_df, activity$interval[1:288])

plot(mean_steps_df2$`activity$interval[1:288]`, mean_steps_df2$mean_steps, type = "l", xlab = "Time interval", ylab = "Average steps taken per time interval across all days", main = "Average daily pattern")

max_steps <- as.integer(max(mean_steps))

```

The maximum number of steps on average occurs at the 835th time interval and it is `r max_steps`

## Task 3 - Imputing missing values

Calculate the total number of missing values (NAs)

``` {r missing NAs count}

activity <- read.csv("activity.csv")

activity_na_count <- activity[rowSums(is.na(activity)),]

nos_na <- nrow(activity_na_count)
```

The total number of missing values in the dataset is `r nos_na`

Apply a strategy to replace missing values - replace missing values with the mean for that time interval

``` {r replace missing values}

library(dplyr)

mean_steps_df2 <-  mean_steps_df2 %>% rename(interval = `activity$interval[1:288]`)

activity <- read.csv("activity.csv")

for (i in 1: 17568){
        if (is.na(activity$steps[i] == TRUE)){
                interval <- activity$interval[i]
                activity$steps[i] <- mean_steps_df2$mean_steps[match(interval, mean_steps_df2$interval)]
        }
        i <- i + 1
        
}

total_steps_nona <- tapply(activity$steps, activity$date, sum)

total_steps_nona <- total_steps_nona[!is.na(total_steps)]

hist(total_steps_nona, col = "red", xlab = "Total number of steps taken each day", main = "Histogram of steps taken each day")

mean_total_steps_nona <- as.integer(mean (total_steps_nona))

median_total_steps_nona <- as.integer(median (total_steps_nona))


```

The mean for the number of steps taken each day with NAs removed is `r mean_total_steps_nona`

The median for the number of steps taken each day with NAs removed is `r median_total_steps_nona`

These values **DO NOT** differ from the estimates of the mean and median in the first part

``` {r compare total steps}

par(mfrow = c(1,2))

hist(total_steps, col = "red", xlab = "Total number of steps taken each day", main = "No NAs")

hist(total_steps_nona, col = "blue", xlab = "Total number of steps taken each day", main = "Imputing NAs")
```

It can be seen from the graphs above that imputing NAs bumps up the frequency for the number of steps in the 10,000-15,000 range but the overall pattern remains the same

## Task 4 - Differences in activity patterns between weekdays and weekends

``` {r activity weekdays or weekends}

library(lubridate)

activity$date <-  ymd(activity$date)

activity$day <- weekdays(activity$date, abbreviate = TRUE)

activity$day_type <- ifelse(weekdays(activity$date) %in% c("Saturday", "Sunday"), "weekend", "weekday")

activity_wekd <- activity %>% filter(day_type == "weekday")

activity_wekend <- activity %>% filter(day_type == "weekend")

par(mfrow = c(2,1))

plot(activity_wekd$interval,activity_wekd$steps, type = "l", xlab = " Interval", ylab = "Number of steps", main = "Weekday")

plot(activity_wekend$interval, activity_wekend$steps, type = "l", xlab = "Interval", ylab = "Number of steps", main = "Weekend")


```









